---
title: "Assignment 2"
format: pdf
author: "Jessica Stow (STWJES003@MYUCT.AC.ZA)"
date: "14 October 2024"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      error = FALSE,
                      include = FALSE,
                      warning = FALSE, 
                      cache = TRUE)
```

# View this report on my GitHub profile!

This report's repository can be viewed on [my GitHub profile](https://github.com/jessicastow/neural-networks).

# Plagiarism declaration

-   I know that plagiarism is wrong.

-   Plagiarism is to use another’s work and pretend that it is one’s own.

-   I have used the required convention for citation and referencing.

-   Each contribution to and quotation in this assignment from the work(s) of other people has been attributed, and has been cited and referenced.

-   This assignment is my own work.

-   I have not allowed, and will not allow, anyone to copy my work with the intention of passing it off as his or her own work.

-   I acknowledge that copying someone else’s assignment or essay, or part of it, is wrong, and declare that this is my own work.

\newpage

# Classification task

```{r}
classification <- read.csv("data/Data-classification.csv")
```



```{r}
str(classification)
unique(classification$Target)
```

# Regression task

```{r}
regression <- read.csv("data/Data-regression.csv")
str(regression)
```

```{r}
regression$target
```


